Hydraulically-vulnerable trees are buffered by deep-water access in a tropical forest community

### 2020-10-21 manuscript.rmd
1. changed chosen ERD model 2. changed depths used for ERD 3. LMA and WSG to Roman 4. P to small p roman; rho to r 5. space

\begin{equation}
G_{s,~t~|~z,~h} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} (K^{*}_{leaf,~s,~i~|~z,~h} + VPD^{*}_{i}) LAI^{*}_{s,~doy~|~i}) + \epsilon_{t~|~z,~h}
(\#eq:M-chosen)
\end{equation}

@@

By varying ‘z’, we estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). Using Eq. \@ref(eq:M-chosen), for a species $s$, we obtained Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations. For each hydrological realization we retained only one depth that corresponded to the least AIC score (the best-fit) [@Chonggang, we could have retained multiple depths with non-differentiable AICs], and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $r^2$ >= 0.5) as a valid set. For a species $s$, we defined effective rooting depth (ERD) as the median of these set of depths, described as :

\begin{equation}
ERD_{s} = \underset{\forall \, h \, \in \, H}{median} \Big\{z,h\,\Big|\, \Big(AIC_{s~|~z,~h} = \underset{\forall \, z \,\in\,Z}{min}(AIC_{s~|~z,~h})\Big) \cap \Big(r_{s~|~z,~h} \ge 0.71\Big)\Big\}
(\#eq:erd)
\end{equation} 
where $H$ $\in$ {1, .., 100}, refer to the 100 hydrological realisations and $Z$ $\in$ {`r paste(soil.depths, collapse=", ")`} m, covering the range of soil depths in ELM-FATES after above parametarision.

@@

#### Leaf Area Index (LAI)
\
To explore the effect of species seasonality in leaf phenology on growth, we assumed species-specific seasonal curves for leaf area index ($LAI$, range 0-1, unitless). For evergreen species we assumed $LAI$ = 1 throughout the year, whereas for non-evergreen species (various deciduous classes), we assumed $LAI$ = 1 during the wet season (DOY = 150 to 300) given the high forest-level leaf area index for this period (not shown), and obtained an estimate of $LAI$ for the rest of the year as follows. We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate, scaled to express daily leaf fall as a fraction of the total leaf fall in a given year and obtained an average leaf fall fraction by DOY estimate across years ($LeafFall$, range 0-1, unitless). Thus for non-evergreen species, we cumulatively deducted $LeafFall$ from $LAI$ = 1 beginning DOY = 300 and obtained a $LAI$ by DOY curve (Fig. S\@ref(fig:leaf-cover)). We obtained species-specific parameters $A$ & $B$ for Eq. \@ref(eq:k-soil) either from data when available, or a model based on $LMA$ or $WSG$ as described in the earlier section, and used K standardized between 0 and 1 within species ($K^*$) in Eq. \@ref(eq:M-chosen).

@@

## Leaf hydraulic vulnerability curves
\
For the leaf vulnerability curve parameters $A$ and $B$ fitted to the observed data for $K_{leaf}$ vs. $\Psi_{leaf}$, $WSG$ and parameter $B$ explained a large proportion of variation in parameter $A$ (Eq. \@ref(eq:a-mod); *Adj. R*^2^ = `r afit.r2`; *P* `r  afit.p`; Fig. \@ref(fig:ab-fit)). $WSG$ and $LMA$ explained a large proportion of variation in parameter $B$ (Eq. \@ref(eq:b-mod); Adj. *R*^2^ = `r bfit.r2`; *P* `r  bfit.p`; also see Fig. S\@ref(fig:obs-plc-curves). Vulnerability curve parameters were predicted for `r sp.n.model` out of `r length(unique(erd.iso$sp))` ERD species without direct data using Eqs. \@ref(eq:a-mod)-\@ref(eq:b-mod) (Table S\@ref(tab:ab-param), Fig.  S\@ref(fig:obs-comm-plc-curves)). 
Fig. \@ref(fig:ab-fit) here.

## Effective rooting depths (ERD)
\
ERD estimated from the model with K~leaf~, LAI and an additive effect of VPD as explanatory variables (Eq. \@ref(eq:M2)) was strongly related to the measured isotopic signal ($\delta^2H$), that is, with highest goodness-of-fit (*R*^2^ = 0.65; *P* = 0.027; Fig. \@ref(fig:erd-iso)). The model with K~leaf~, but without LAI and an additive effect of VPD showed no species with sufficient species-level goodness-of-fit. ERD from all the other models showed a positive, but non-significant relationship with $\delta^2H$ at $\alpha$ = 0.05 (Fig. S\@ref(fig:other-erd-mods)). For the rest of the analyses we therefore used ERD estimates from the best-fit model (Eq. \@ref(eq:M-chosen)). Notably, two species, *L. seemannii* and *T. tuberculata*, that @Meinzer:1999 found to have xylem sap $\delta^2H$ values between soil water and groundwater, suggesting that these species sourced most of the water from depths > 1 m, were also identified by our model with ERD > 1 m (2.9 m for both species; Fig. \@ref(fig:erd-iso)).

@@

  For the subset of 8 species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max,~stem}$, Spearmans' *r* = `r cor.k`, *P* = `r p.k` at $\alpha$ = 0.05; Fig. \@ref(fig:erd-hyd)A), less negative leaf $\Psi_{tlp}$ (*r* = `r cor.tlp`, *P* = `r p.tlp`; Fig. \@ref(fig:erd-hyd)B), less negative $\Psi_{88,~stem}$ and thus greater vulnerability to embolism from cavitation (*r* = `r cor.stem.88`, *P* = `r p.stem.88`; Fig. \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *r* = `r cor.hsm`, *P* = `r p.hsm`; Fig. \@ref(fig:erd-hyd)D).

@@

## Effective rooting depths, experienced drought and mortality
\
Species with deeper ERD showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *P* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig. \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower ERD had significantly higher mortality anomalies than deeper ERD species (Fig. S\@ref(fig:erd-mort-census)). Analysis of time spent under significant soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD, that is the shallower the ERD, the greater proportion of days $\Psi_{ERD}$ < $\Psi_{crit}$ (Fig. \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yr period for which soil water potential $\Psi_{soil,~z}$ was modeled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`;  median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).

@@

\
Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018 in which species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, extreme drought. Their hydrological model revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity. In contrast, our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi_{soil}$ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa). The difference in our results are likely attributable to site differences in the precipitation regime. In their dry forest site (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016), and thus at BCI deep soil layers are effectively recharged annually (Fig. S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018). The variation in mortality explained by ERD in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality [@McDowell.2018].

@@

### Future directions for model development
\
The trade-offs that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests. The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high $K_{max,~stem}$) and cavitation-resistance (low $\Psi_{50,~stem}$ or $\Psi_{88,~stem}$) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016]. We report such a trade-off (n = 26 species) at our study site stem level (). We also found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown). WSG alone explained `r round(a.wsg.r2*100,0)` % of variation in parameter $A$ ($A = `r a.wsg.cf[1]` `r a.wsg.cf[2]`B + `r a.wsg.cf[3]`B^2 + error$; *P* `r  a.wsg.p`) and `r round(b.wsg.r2*100,0)` % of variation in $B$ ($B = `r b.wsg.cf[1]` `r b.wsg.cf[2]`B + `r b.wsg.cf[3]`B^2 + error$; *P* `r  b.wsg.p`). Parameter $B$ alone explained `r round(a.b.r2*100,0)` % of variation in $A$ (*P* `r  a.b.p`), while $LMA$ alone was weakly related to $A$ and was not related to $B$ (not shown). The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.

@@

Figure \@ref(fig:ab-fit): Parameters $A$ (panel A) & $B$ (panel B) for the leaf vulnerability curves that were fitted to observed data on $K_{leaf}$ vs. $\\Psi_{leaf}$ for `r length(sp.exp.param$sp)` species (Eq. \@ref(eq:vc-exp)) vs. those that were predicted from model based on trait-proxies. (Eqs. \@ref(eq:a-mod)-\@ref(eq:b-mod)). Goodness-of-fit (*R*^2^) and *P*-values are given in inset.

@@

Figure \@ref(fig:erd-iso): Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for `r nrow(cor.data)` canopy species found on Barro Colorado Island (Panama). Each circle is color coded by the Spearman's $\\rho$ with red shades for positive $\\rho$ values, while blue shades for negative values, respectively. P-values are given only when they are $\\ge$ 0.05. Units are as follows: ERD in m, K~max~ in mmol m^-1^ s^-1^ MPa^-1^), all variables with $\\Psi$ in MPa.

Figure \@ref(fig:erd-mort): Mortality rate (mean $\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r length(erd.mrate.sp)` canopy species found on Barro Colorado Island. *R*^2^ and *P*-values are given in panel insets.

@@

Among the `r length(sp.exp.param$sp)` species with observed data on ($K_{leaf}$ vs. $\Psi_{leaf}$) used to build the community-level models of leaf vulnerability curves (Eqs. \@ref(eq:b-mod)-\@ref(eq:a-mod)), two lacked data on WSG and five on LMA~lamina~. We filled three gaps in LMA using predictions from a strong linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{LMA} = `r disc.cf[1]` + `r disc.cf[2]`LMA_{disc}$; Adj. *R*^2^ = `r disc.r2`; *P* `r disc.p`, *n* = `r disc.n`). We filled the two remaining gaps in LMA using predictions from a linear relationship between LMA and LMA~lamina~, as measured during the smaller, $K_{leaf}$ campaign ($\overline{LMA} = `r lma.cf[1]` + `r lma.cf[2]`LMA_{lamina}$; Adj. *R*^2^ = `r lma.r2`; *P* `r lma.p`, *n* = `r lma.n`). We chose not to fill the two data gaps in WSG as available data were poor predictors of WSG.

@@

\begin{equation}
G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} VPD_{i}^{*}) + \epsilon_{t~|~z}
(\#eq:M2)
\end{equation}

\begin{equation}
G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} + VPD_{i}^{*}) + \epsilon_{t~|~z}
(\#eq:M3)
\end{equation}

@@

```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths (z): 0.1, 0.6 and 1 m (vertical panels). Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\\Psi_{soil,~z}$) for a given DOY and depth z are shown (same across census intervals). $\\Psi_{soil,~z}$ by DOY for a year (colored line) is only shown if for some part of that year $\\Psi_{soil,~z}$ declined below the 2.5 percentile distribution."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q2.5_by_depth.jpeg"))
```

```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_{leaf}$ vs. $\\Psi_{leaf}$ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D). Each line represents a species, color coded by $WSG$ (A, C), or $LMA$ (B, D), with blue tones indicating smaller values of $WSG$ or $LMA$ and red tones indicating larger values.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
```

@@

```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\delta^2H$ for each of the ERD models tested (panels). For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's $\\rho$ > 0 and *R*^2^ > 0.5 during ERD parameterisation. The number of species with an ERD estimate as well as $\\delta^2H$. or each model are shown in inset, along with the goodness-of-fit (*R*^2^) and *P*-values for a linear relationship between ERD and $\\delta^2H$. ERD models in panels A-E correspond to Eq. \\@ref(eq:M1)-\\@ref(eq:M6), respectively.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_sp_color_Meinzer.jpeg"))
```

>>>>

\begin{equation}
G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} VPD_{i}^{*}) + \epsilon_{t~|~z}
(\#eq:M-chosen)
\end{equation}

@@

By varying ‘z’, we estimated species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). Using Eq. \@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations. For each hydrological realization we retained only one depth that corresponded to the least AIC score (the best-fit), and within those depths retained the ones that had Pearson's $r$ >= 0.71 (or, $r^2$ >= 0.5) as a valid set. For a species $s$, we defined effective rooting depth (ERD) as the median of these set of depths, described as :

\begin{equation}
ERD_{s} = \underset{\forall \, h \, \in \, H}{median} \Big\{z,h\,\Big|\, \Big(Likelihood_{s~|~z,~h} = \underset{\forall \, z \,\in\,Z}{maximum}(Likelihood_{s~|~z,~h})\Big) \cap \Big(r_{s~|~z,~h} \ge 0.71\Big)\Big\}
(\#eq:erd)
\end{equation} 
where $H$ $\in$ {1, .., 100}, refer to the 100 hydrological realisations and $Z$ $\in$ 0.5, {`r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m, covering the range of soil depths in ELM-FATES after above parametarision [@Chonggang, Depths 0.21, 0.37 and 0.62 were all labelled as 0.5 m for model evaluation. How do I explain that? Do they need to be retained as their original depths?].

@@

#### Leaf Area Index (LAI)
\
To explore the effect of species seasonality in leaf phenology on growth, we assumed species-specific seasonal curves for leaf area index ($LAI$, range 0-1, unitless). For evergreen species we assumed $LAI$ = 1 throughout the year, whereas for non-evergreen species (various deciduous classes), we assumed $LAI$ = 1 during the wet season (DOY = 150 to 300) given the high forest-level leaf area index for this period (not shown), and obtained an estimate of $LAI$ for the rest of the year as follows. We interpolated species-specific weekly leaf-fall data collected at two similar sites in BCI (1985-2020) to a daily estimate, scaled to express daily leaf fall as a fraction of the total leaf fall in a given year and obtained an average leaf fall fraction by DOY estimate across years ($LeafFall$, range 0-1, unitless). Thus for non-evergreen species, we cumulatively deducted $LeafFall$ from $LAI$ = 1 beginning DOY = 300 and obtained a $LAI$ by DOY curve (Fig. S\@ref(fig:leaf-cover)). We obtained species-specific parameters $A$ & $B$ for Eq. \@ref(eq:k-soil) either from data when available, or a model based on LMA or WSG as described in the earlier section, and used K standardized between 0 and 1 within species ($K^*$) in Eq. \@ref(eq:M-chosen).

@@

## Leaf hydraulic vulnerability curves
\
For the leaf vulnerability curve parameters $A$ and $B$ fitted to the observed data for $K_{leaf}$ vs. $\Psi_{leaf}$, WSG and parameter $B$ explained a large proportion of variation in parameter $A$ (Eq. \@ref(eq:a-mod); *Adj. R*^2^ = `r afit.r2`; *p* `r  afit.p`; Fig. \@ref(fig:ab-fit)). WSG and LMA explained a large proportion of variation in parameter $B$ (Eq. \@ref(eq:b-mod); Adj. *R*^2^ = `r bfit.r2`; *p* `r  bfit.p`; also see Fig. S\@ref(fig:obs-plc-curves). Vulnerability curve parameters were predicted for `r sp.n.model` out of `r length(unique(erd.iso$sp))` ERD species without direct data using Eqs. \@ref(eq:a-mod)-\@ref(eq:b-mod) (Table S\@ref(tab:ab-param), Fig.  S\@ref(fig:obs-comm-plc-curves)). 
Fig. \@ref(fig:ab-fit) here.

## Effective rooting depths (ERD)
\
ERD estimated from the model with K~leaf~, LAI and an additive effect of VPD as explanatory variables (Eq. \@ref(eq:M2)) was strongly related to the measured isotopic signal ($\delta^2H$), that is, with highest goodness-of-fit (*R*^2^ = 0.65; *p* = 0.027; Fig. \@ref(fig:erd-iso)). The model with K~leaf~, but without LAI and an additive effect of VPD showed no species with sufficient species-level goodness-of-fit. ERD from all the other models showed a positive, but non-significant relationship with $\delta^2H$ at $\alpha$ = 0.05 (Fig. S\@ref(fig:other-erd-mods)). For the rest of the analyses we therefore used ERD estimates from the best-fit model (Eq. \@ref(eq:M-chosen)). Notably, two species, *L. seemannii* and *T. tuberculata*, that @Meinzer:1999 found to have xylem sap $\delta^2H$ values between soil water and groundwater, suggesting that these species sourced most of the water from depths > 1 m, were also identified by our model with ERD > 1 m (2.9 m for both species; Fig. \@ref(fig:erd-iso)).

@@

  For the subset of 8 species with ERD for which hydraulic traits data were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max,~stem}$, Spearmans' *r* = `r cor.k`, *p* = `r p.k` at $\alpha$ = 0.05; Fig. \@ref(fig:erd-hyd)A), less negative leaf $\Psi_{tlp}$ (*r* = `r cor.tlp`, *p* = `r p.tlp`; Fig. \@ref(fig:erd-hyd)B), less negative $\Psi_{88,~stem}$ and thus greater vulnerability to embolism from cavitation (*r* = `r cor.stem.88`, *p* = `r p.stem.88`; Fig. \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *r* = `r cor.hsm`, *p* = `r p.hsm`; Fig. \@ref(fig:erd-hyd)D).

@@

## Effective rooting depths, experienced drought and mortality
\
Species with deeper ERD showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *p* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig. \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower ERD had significantly higher mortality anomalies than deeper ERD species (Fig. S\@ref(fig:erd-mort-census)). Analysis of time spent under significant soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD, that is the shallower the ERD, the greater proportion of days $\Psi_{ERD}$ < $\Psi_{crit}$ (Fig. \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yr period for which soil water potential $\Psi_{soil,~z}$ was modeled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`;  median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).

@@

\
Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018 in which species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, extreme drought. Their hydrological model revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity. In contrast, our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period, $\Psi_{soil}$ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa). The difference in our results are likely attributable to site differences in the precipitation regime. In their dry forest site (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016), and thus at BCI deep soil layers are effectively recharged annually (Fig. S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018). The variation in mortality explained by ERD in our study is low (*R*^2^ = 0.11, *p* = 0.075), and other causes may explain observed mortality [@McDowell.2018].

@@

### Future directions for model development
\
The trade-offs that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests. The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high $K_{max,~stem}$) and cavitation-resistance (low $\Psi_{50,~stem}$ or $\Psi_{88,~stem}$) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016]. We report such a trade-off (n = 26 species) at our study site stem level (). We also found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown). WSG alone explained `r round(a.wsg.r2*100,0)` % of variation in parameter $A$ ($A = `r a.wsg.cf[1]` `r a.wsg.cf[2]`B + `r a.wsg.cf[3]`B^2 + error$; *p* `r  a.wsg.p`) and `r round(b.wsg.r2*100,0)` % of variation in $B$ ($B = `r b.wsg.cf[1]` `r b.wsg.cf[2]`B + `r b.wsg.cf[3]`B^2 + error$; *p* `r  b.wsg.p`). Parameter $B$ alone explained `r round(a.b.r2*100,0)` % of variation in $A$ (*p* `r  a.b.p`), while LMA alone was weakly related to $A$ and was not related to $B$ (not shown). The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.

@@

Figure \@ref(fig:ab-fit): Parameters $A$ (panel A) & $B$ (panel B) for the leaf vulnerability curves that were fitted to observed data on $K_{leaf}$ vs. $\\Psi_{leaf}$ for `r length(sp.exp.param$sp)` species (Eq. \@ref(eq:vc-exp)) vs. those that were predicted from model based on trait-proxies. (Eqs. \@ref(eq:a-mod)-\@ref(eq:b-mod)). Goodness-of-fit (*R*^2^) and *p*-values are given in inset.

@@

Figure \@ref(fig:erd-iso): Matrix for pairwise correlations (circles) among Effective Rooting Depth (ERD) and hydraulic properties for `r nrow(cor.data)` canopy species found on Barro Colorado Island (Panama). Each circle is color coded by the Spearman's *r* with red shades for positive *r* values, while blue shades for negative values, respectively. *p*-values are given only when they are $\\ge$ 0.05. Units are as follows: ERD in m, K~max~ in mmol m^-1^ s^-1^ MPa^-1^), all variables with $\\Psi$ in MPa.

Figure \@ref(fig:erd-mort): Mortality rate (mean $\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) for `r length(erd.mrate.sp)` canopy species found on Barro Colorado Island. *R*^2^ and *p*-values are given in panel insets.

@@

Among the `r length(sp.exp.param$sp)` species with observed data on ($K_{leaf}$ vs. $\Psi_{leaf}$) used to build the community-level models of leaf vulnerability curves (Eqs. \@ref(eq:b-mod)-\@ref(eq:a-mod)), two lacked data on WSG and five on LMA~lamina~. We filled three gaps in LMA using predictions from a strong linear relationship between LMA and LMA~disc~, the latter defined as the mean leaf mass per unit area measured for a 1.483 cm^2^ leaf disc taken to avoid veins (g m^-2^) for leaves receiving direct sunlight ($\overline{LMA} = `r disc.cf[1]` + `r disc.cf[2]`LMA_{disc}$; Adj. *R*^2^ = `r disc.r2`; *p* `r disc.p`, *n* = `r disc.n`). We filled the two remaining gaps in LMA using predictions from a linear relationship between LMA and LMA~lamina~, as measured during the smaller, $K_{leaf}$ campaign ($\overline{LMA} = `r lma.cf[1]` + `r lma.cf[2]`LMA_{lamina}$; Adj. *R*^2^ = `r lma.r2`; *p* `r lma.p`, *n* = `r lma.n`). We chose not to fill the two data gaps in WSG as available data were poor predictors of WSG.

@@

\begin{equation}
G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} + VPD_{i}^{*}) + \epsilon_{t~|~z}
(\#eq:M2)
\end{equation}

\begin{equation}
G_{t~|~z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{leaf,~i~|~z}^{*} VPD_{i}^{*}) + \epsilon_{t~|~z}
(\#eq:M3)
\end{equation}

@@

```{r swp-droughts, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "100%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-droughts}Occurrence of extreme soil droughts by census intervals (horizontal panels) at three representative depths (z): 0.1, 0.6 and 1 m (vertical panels). Mean (black lines) and lower half of 95% distribution (grey areas) of soil water potential ($\\Psi_{soil,~z}$) for a given DOY and depth z are shown (same across census intervals). $\\Psi_{soil,~z}$ by DOY for a year (colored line) is only shown if for some part of that year $\\Psi_{soil,~z}$ declined below the 2.5 percentile distribution."}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi_model_daily_bestfit_params.top.few_CI_full_interval_panels_climatology_over_study_period_q5_by_depth.jpeg"))
```

```{r obs-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "50%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-obs-plc-curves}Vulnerability curves fitted to observed $K_{leaf}$ vs. $\\Psi_{leaf}$ data on ", length(sp.exp.param$sp)," tree species from BCI (A & B) with equation \\@ref(eq:vc-exp)--in absolute terms (A, B), as well as, in terms of loss of conductivity (%) (C, D). Each line represents a species, color coded by WSG (A, C), or LMA (B, D), with blue tones indicating smaller values of WSG or LMA and red tones indicating larger values.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/plot.obs.plc.jpeg"))
```

@@

```{r other-erd-mods, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-other-erd-mods}Effective Rooting Depth versus $\\delta^2H$ for each of the ERD models tested (panels). For each ERD model only those species have an estimate of ERD that have passed the test of Pearson's *r* > 0 and *R*^2^ > 0.5 during ERD parameterisation. The number of species with an ERD estimate as well as $\\delta^2H$. or each model are shown in inset, along with the goodness-of-fit (*R*^2^) and *p*-values for a linear relationship between ERD and $\\delta^2H$. ERD models in panels A-E correspond to Eq. \\@ref(eq:M1)-\\@ref(eq:M6), respectively.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_sp_color_Meinzer.jpeg"))
```
